State of the Practice of Lattice Boltzmann Method Software Development

### 2021-06-23 StateOfPractice/Peter-Notes/Report/Report.tex
LBM 5.4 small edits

Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles.

Several development process obstacles were identified. A developer of pyLBM noted that their small development team has a lack of time to implement new features. Small development teams are common for LBM software packages. Team members are almost always part of the same institute or already know each other from other projects. External contributions are rare despite many of the projects accepting them.

A lack of software development experience was noted by the developer of TCLB. Many of the team members on their project are domain experts and there can be a steep learning curve before these team members contribute good quality source code. It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions. The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software. The importance of funding for scientific software has been discussed in \cite{gewaltig2012quality}, which notes that ``software tools are developed and maintained only for as long as there is explicit or implicit funding''. The develop also commented that there are no journals that publish such scientific software source code. However, some work has been done to address such concerns in \citep{smith2016software}, which presents a set of software citation principles and discusses ``how they could be used to implement software citation in the scholarly community'' \citep{katz2019software}. Another raised concern was that there are no organizations helping with the development of good quality software; but some do exist, including \href{https://bssw.io/}{Better Scientific Software (BSSw)}, \href{https://www.software.ac.uk/}{Software Sustainability Institute}, and \href{https://software-carpentry.org/}{Software Carpentry}. Some SCS developers may not be familiar with these organizations.

Scientific software is often developed in-house by the very researchers that temporarily use it in their own research. Empirical studies of such ``professional end-user development'' of SCS is noted in \citep{segal2007end}. This kind of software has a defined user and purpose, and often does not meet the standards that would be required by external users. It has been categorized as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite often being made freely available, ``it is not always clear that it is sufficiently mature in terms of domain coverage, validity, documentation or usability, to be useful to other researchers''. It is less common for such ad hoc scientific software to become ``user-ready software'', which ``is not only research-ready, but should have most of the attributes commonly expected of commercial software products including broadness of scope, robustness, demonstrable correctness and adequate documentation'' \citep{gewaltig2012quality}. As software becomes user-ready, it can become commercialized and closed-source.

Documentation of LBM software is noted as being important, but its quality could be improved. There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team. Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory. Users would be well advised to already be familiar with these topics, or they should spend significant time referencing theory resources. The documentation instead generally focuses on explaining how to use the software. It is of course not feasible for package documentation to address the underlying physics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory. Sometimes, frequently asked questions about the underlying theory are answered in the documentation. OpenLB has an artifact for such questions.

>>>>

Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles. Several development process obstacles were identified.

A developer of pyLBM noted that their small development team has a lack of time to implement new features. Small development teams are common for LBM software packages. Team members are almost always part of the same institute or already know each other from other projects. External contributions are rare despite many of the projects accepting them. Aside from on-boarding new developers, time constraints could be mitigated by increasing developer efficiency, which could be addressed in several ways, including by improving the quality of documentation, or incorporating automatic code generation.

A lack of software development experience was noted by the developer of TCLB. Many of the team members on their project are domain experts and there can be a steep learning curve before these team members contribute good quality source code. It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions.

The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software. The importance of funding for scientific software has been discussed in \cite{gewaltig2012quality}, which notes that ``software tools are developed and maintained only for as long as there is explicit or implicit funding''. The developer further commented that there are no journals that publish such scientific software source code. However, there are ways to get such source code cited. Work has been done to address this in \citep{smith2016software}, which presents a set of software citation principles and discusses ``how they could be used to implement software citation in the scholarly community'' \citep{katz2019software}.

Another raised concern was that there are no organizations helping with the development of good quality software; but some do exist, including \href{https://bssw.io/}{Better Scientific Software (BSSw)}, \href{https://www.software.ac.uk/}{Software Sustainability Institute}, and \href{https://software-carpentry.org/}{Software Carpentry}. Some SCS developers may not be familiar with these organizations.

Scientific software is often developed in-house by the very researchers that temporarily use it in their own research. Empirical studies of such ``professional end-user development'' of SCS is noted in \citep{segal2007end}. This kind of software has a defined user and purpose, and often does not meet the standards that would be required by external users. It has been categorized as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite often being made freely available, ``it is not always clear that it is sufficiently mature in terms of domain coverage, validity, documentation or usability, to be useful to other researchers''. It is less common for such ad hoc scientific software to become ``user-ready software'', which ``is not only research-ready, but should have most of the attributes commonly expected of commercial software products including broadness of scope, robustness, demonstrable correctness and adequate documentation'' \citep{gewaltig2012quality}. As software becomes user-ready, it can become commercialized and closed-source. Documentation of LBM software is noted as being important, but its quality could be improved. As already noted, there is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team. Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory. Users would be well advised to already be familiar with these topics, or they should spend significant time referencing theory resources. The documentation instead generally focuses on explaining how to use the software. It is of course not feasible for package documentation to address the underlying physics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory. Sometimes, frequently asked questions about the underlying theory are answered in the documentation. OpenLB has an artifact for such questions.
